---
title: "MA7442 Assessment 1 - Task 1"
format: html
---

# AIM :

-   The aim of this study is to Examining the **socioeconomic**, **behavioral**, and **demographic characteristics** linked to diabetes prevalence in the study population as well as identifying important predictors that might influence variation in **diabetes status** are the goals of this research.

# Objectives :

-   To examine the distribution of diabetes status within the dataset.

-   To explore the relationships between diabetes and demographic variables like **gender** and **age**.

-   To evaluate the connection between **biological risk factors**, specifically **BMI**, and **diabetes**.

-   To explore into the relationship between the prevalence of diabetes and lifestyle choices (such as **diet** and **alcohol** use).

-   To explore the relationship between diabetes risk and socioeconomic factors, such as **income level**.

-   To employ multivariable modeling to identify the variables that, after adjusting for confounding variables, independently predict diabetes. (**Task 2**)

## Dataset

### Read in Package and data

```{r}
#| warning: false

library(tidyverse)
library(ggplot2)
library(dplyr)
library(reshape2)

```

```{r}
data = read.csv("data/dataset1.csv")

```

## Overview

The dataset chosen for this research contains healthcare statistics and lifestyle survey information from the US as well as diabetic state.

-   It contains **253,680** rows and **22** columns with **zero** missing values.

-   It takes into account factors like **age**, **body mass index (BMI)**, **alcohol** and **fruit intake**, **income** and **diabetes**.

-   Most of the values in feature classified as **0's** and **1's** which means **no** and **yes**.


```{r}
head(data)
```


```{r}
dim(data)
```


```{r}
data[data == "" | data == "?"] <- NA
colSums(is.na(data))
```

```{r}
table(data$Diabetes_012)
table(data$HighBP)
table(data$HighChol)
table(data$Smoker)

```

-   The majority of people (**n = 213,703**) do not have diabetes. Prediabetes cases make up the smallest group (**n=4,631**) and only **35,346** cases are diabetes. it will affect classification performance and This **imbalance** should be taken into account in later modeling.

-   With **112,423** people identified as smokers and **141,257** people are non-smokers.

# Data Preprocessing

```{r}
data$Age <- factor(data$Age,
                   levels = 1:13,
                   labels = c("18-24","25-29","30-34","35-39","40-44","45-49","50-54","55-59","60-64","65-69","70-74","75-79","80+"),
                   ordered = TRUE
)

```

-   Initially, the age variable was numerically coded from level 1 to 13 but those number are age groups and not real ages,So convert it into a **categorical variable**.

```{r}
data$Education <- factor(data$Education,
                         levels = 1:6,
                         labels = c("Never attended school or
only kindergarten", "Grades 1-8","Grades 9-11", "Grade 12" ,"College 1 year-3 years", "College graduate" ),
ordered = TRUE

                         
  
)
```

```{r}
data$GenHlth <- factor(data$GenHlth,
                       levels = 1:5,
                       labels = c("excellent", "very good","good","fair","poor"),
                       ordered = TRUE
  
)
```

```{r}
data$Income <- factor(data$Income,
  levels = 1:8,
  labels = c("<$10k", "$10k-15k", "$15k-20K","$20k-25k", "$25-35K", "$35k-40k","$40-75k",  "+$75k"),
  ordered = TRUE
)
```

```{r}
table(data$Age)
table(data$Education)
table(data$GenHlth)
table(data$Income)
table(data$Sex)
```

-   The middle and older age groups make up the majority of age distribution with 60-64 and 65-69 age groups shows highest frequencies.

-   Majority are college graduates - 107,325

-   Most of the report **Very Good** or **Good** in General health level .

-   There are *90,385* people who earn more than *\$75k*.

-   There are **141,974** Female count in data and **111,706** male count in data. This shows it's relatively balanced by sex.

```{r}
summary(data)
```

```{r}
table(data$Diabetes_012)
prop.table(table(data$Diabetes_012))
```

# Data Visualization

## Distribution of age

```{r}
ggplot(data, aes(x= Age))+
  geom_bar(fill= "darkgreen")+
  labs(title =  "Distribution of Age",
         x = "Age")

```

```{r}
ggplot(data, aes(x= factor(Diabetes_012), fill = factor(Diabetes_012)))+
  geom_bar(position = "dodge")+
  labs(title = "Distribution of Diabetes",
       x= "Diabetes",
       y= "frequency")
```

-   The bar Graph distribution shows frequency of diabetes. There is a huge amount of data which as no diabetes and less as diabetes.This is imbalanced.

## Distribution of BMI

```{r}

ggplot(data,aes(x= BMI))+
  geom_histogram(aes(y= ..density..),
                 bandwith=2,fill= "lightgreen",color= "orange")+
  stat_function(fun= dnorm,
          args= list(mean= mean(data$BMI),
                     sd= sd(data$BMI)),
    color = "red", linewidth=1)+
  labs(
    title = "Boby mass index",
    x = "BMI", 
    y = "Density"
  )
```

-   The Histogram shows **Body Mass Index** (BMI) values with density curve.The distribution is **right-skewed**(positively skewed).Tells most values are in lower end and less values in tail part.

```{r}
ggplot(data, aes(x = factor(Diabetes_012),y = BMI, fill= factor(Diabetes_012)))+
  geom_boxplot()+
  labs(
    title =  "BMI Distribution by Diabetes",
    x = "Diabetes",
    y = "BMI"
  )
        
```

-   We can see lot of outliers in Boxplot which we remove in task2.

## Age vs Diabetes

```{r}
ggplot(data,aes(x= Age, fill = factor(Diabetes_012)))+
         geom_bar(position = "dodge")+
         coord_flip()+
         labs(
           title = "Diabetes by different ages",
           x= "Age"
         )
```

-   In this bar graph Diabetes is **strongly correlated** to age. The percentage of people with diabetes rises steadily as people age, with the **highest** frequency seen in those between the ages of **65-69**. Diabetes rates are significantly lower in younger age groups. This trend aligns with the established notion that aging is a significant risk factor for diabetes.

```{r}
age_dia <- data |>
  group_by(Age) |>
  summarise(
    diabete_mean = mean(Diabetes_012==2)
  )
  
ggplot(age_dia,aes(x= Age,y= diabete_mean,group=1,color= "diabete_mean"))+
  geom_line()+
  geom_point()+
  labs(
    title = "Diabete frequency between different Age group",
    x = "Age",
    y = "Mean of diabetes"
  )+
  theme_minimal()+
  theme(axis.text.x = element_text(angle = 45,hjust = 1))

```

-   This line plot show increase in diabetes rate along age group and age_group **70-74** has **highest** chance of diabetes risk.

## Income vs Sex

```{r}
ggplot(data, aes(x= Income,fill = factor(Sex)))+
  geom_bar(position = "dodge")+
  labs(
    title = "Income difference between Sex",
    x = "Income"
  )+
  theme_minimal()
  
```

## Income vs Diabetes

```{r}
inc_dia <- data |>
  group_by(Income) |>
  summarise(
    diabetes_prob = mean(Diabetes_012 == 2))

ggplot(inc_dia,aes(x= Income, y = diabetes_prob,group=1,color= "diabetes_prob"))+
  geom_line()+
  geom_point()+
  labs(
    title = "Income va Diabetes",
    x = "income",
    y = "Income Proportion with diabetes",
    fill= "Diabetes"
  )
```

-   This Line graph shows the lower income have **higher** diabetes frequency,This may reflect because disparities in healthcare access, lifestyle and diet.

## Diabetes vs Sex

```{r}
ggplot(data,aes(x=Diabetes_012, fill = factor(Sex)))+
  geom_bar(position = "dodge")+
  labs(
    title = "Diabetes between different Sex",
    x = "Diabetes"
  )+
  theme_minimal()

```

# Health Behavior Vs Diabetes

## Smoking Vs Diabetes

```{r}
ggplot(data,aes(x= factor(Smoker),fill= factor(Diabetes_012)))+
  geom_bar(position = "dodge")+
  labs(
    title = "Smoking Vs Diabetes",
    x = "Somker 1 and Non-somker 0",
    fill= "Diabetes"
  )
  

```

-   Diabetes and smoking status have a **weak** correlation. Although there are some differences between smokers and non-smokers, these seem to be less significant than those related to age and BMI.

## Heavy alcohol consumer Vs Diabetes

```{r}
ggplot(data,aes(x= factor(HvyAlcoholConsump), fill = factor(Diabetes_012)))+
  geom_bar(position = "dodge")+
  
  labs(
    title = "Alcoholconsumption Status Vs Diabetes",
    x = "Heavy Alcohol Consumption",
    fill = "Diabetes"
    
  )
```

-   While heavy drinkers have a **lower** percentage of diabetes than low-drinkers, this is probably due to influencing variables, especially age, rather than a protective effect of alcohol use.

```{r}
ggplot(data, aes(x= factor(HvyAlcoholConsump),fill= factor(Diabetes_012)))+
  geom_bar(position = "fill")+
  facet_wrap(~Age)+
  coord_flip()+
  labs(
    title = "Heavy Alcohol Consumption Vs Diabetes by Age Group",
    x = "Heavy Alcohol Consumption",
    y = "frequency",
    fill= "Diabetes"
  )
```

## More Fruits Consumer Vs Diabetes

```{r}
ggplot(data, aes(x= factor(Fruits),fill = factor(Diabetes_012)))+
  geom_bar(position = "dodge")+
  labs(
    title = "Fruit Consumption Vs Diabetes",
    x = "Fruit Consumption 1 or more"
  )
```

-   While more fruits consumption have a **higher** percentage of diabetes than low- fruits consumption, this is probably due to influencing variables, especially age, rather than a protective effect of fruits consume.

```{r}
ggplot(data, aes(x=factor(Veggies),fill= factor(Diabetes_012)))+
  geom_bar(position = "dodge")+
  labs(
    title = "Veggies consumption Vs Diabetes",
    x = "veggies"
  )

```

-   While more veggies consumption have a **higher** percentage of diabetes than low- veggies consumption, this is probably due to influencing variables, especially age, rather than a protective effect of veggies consume.

```{r}
numeric_data <- data |>
  select(where(is.numeric))

cor_matrix <- cor(numeric_data)
melted_corr <- melt(cor_matrix)

ggplot(melted_corr, aes(x = Var1, y= Var2, fill = value))+
  geom_tile()+
  scale_fill_gradient2(low = "blue",high = "red",mid = "white",midpoint = 0)+
  
  theme_minimal()+
  theme(axis.text.x = element_text(angle = 45,hjust = 1))
  
```

-   The **heatmap** shows the correlation between all numeric variables in the dataset. Strong positive correlation is indicated by **red**, negative correlation is indicated by **blue**, and weak or no correlation in **white**.

# Developed Research Questions

## 1) Which health-related factors and lifestyle are taking with likelihood of diabetes after accounting for socioeconomic characteristics and demographic?

## 2) What effects do age and socioeconomic status have on the association between diabetes status and lifestyle choices?

## 3) Do people with various diabetes statuses exhibit recognizable patterns in their health behaviors and access to healthcare?


```{r}
write.csv(data, "cleaned_diabetes_data.csv", row.names = FALSE)
```

